---
title: "Covid91 vaccine study Final2021"
author: "Put your name here!"
date: "put the date here"
output: 
  html_document:
    number_sections: true
    toc: true

---

```{r include=FALSE,echo=FALSE}
require(tidyverse)
require(tigerstats)
require(forcats)
OldData <-read.csv(file="covid91.csv",header=TRUE)
#recode to remove the annoying "X" from the dataframe
FinalData <- OldData %>% mutate(infected = recode(infected, aids = "covid", none = "OK" )) %>% select(-one_of("X"))
FinalData$treatment <-as.factor(FinalData$treatment)
FinalData$sex <-as.factor(FinalData$sex)
FinalData$infected <-as.factor(FinalData$infected)
FinalData$LGBTQ <-as.factor(FinalData$LGBTQ)
FinalData$DrugUser <-as.factor(FinalData$DrugUser)
FinalData$treatment<-relevel(FinalData$treatment,ref="placebo")
```

# Introduction

A new  preventative vaccine (covid91) is under test in clinical trials. Given the nature of this vaccine it is suspected that the drug may exhibit different effectiveness across the variety of subpopulations tracked. Consider population subgroups consisting of Males, Females, just the drug users, and just those in the LGBTQ community. Write a complete report that separately considers this drug for each subgroup and your recommendations as to whether people in the subgroup should take the vaccine. Structure your report in four sections each devoted to one of the four subgroups, use null and alternative hypothesis statements and correct statistical verbage when referencing p-values in conclusions.

# Methods

```{r}
Males<-filter(FinalData,sex=="guy")
Females<-filter(FinalData,sex=="gal")
Gay<-filter(FinalData,LGBTQ=="gay")
UserDrugs<-filter(FinalData,DrugUser=="yes")
```

# Overall Effectiveness 
```{r}
barchartGC(~infected + treatment,data=FinalData, main="COVID infections by treatment type",xlab="Infection")
barchartGC(~infected + treatment,data=FinalData, type="percent",main="Percent COVID Infections by treatment type", xlab="Infection")

```
  
Overall, most of the study patients did not get COVID. Of those that were diagnosed with COVID, a greater percent of them received the placebo. Of those that were not diagnosed with COVID, a slightly larger percent of them received the placebo.  

```{r}
OverallTable <- xtabs(~infected + treatment, data=FinalData)
rowPerc(OverallTable)
colPerc(OverallTable)
```
  
Of the study patients who were diagnosed with COVID, 58% of them received the placebo and 42% received the vaccine drug. Of those that were not diagnosed with COVID, 51% received the placebo and 49% received the vaccine drug.
  
```{r}
chisq.test(OverallTable)
chisqtestGC(OverallTable)
fisher.test(OverallTable)
```
  
Due to the small amount of participants who were diagnosed with COVID, the Fisher Exact Test is a better statistical test to interpret for this study. With a small p-value of 0.00001223, we can reject the null hypothesis. This indicates that there is a significant difference in COVID infection rate between those those who receive the treatment and those that receive the placebo. The odds of getting COVID is 1.3418 times greater for those that received the placebo rather than the vaccine.
Overall, this drug is between 53.99% and 60.56% effective.
  
# Effectivness for Males

## Results

### Graphical Descriptive Results

```{r}
barchartGC(~infected + treatment,data=Males, main= "COVID Infections of males by treatment type", xlab="Infection")
barchartGC(~infected + treatment,data=Males, type="percent", main="Percent COVID Infections of males by treatment type", xlab="Infection")
```
  
Of males that were diagnosed with COVID, more of them took the vaccine. Of the males that were not diagnosed with COVID, a slightly larger amount took the placebo.  
    
### Numerical Descriptive Results

```{r}
MaleTable <- xtabs(~infected + treatment, data=Males)
rowPerc(MaleTable)
colPerc(MaleTable)
```
  
Of males that were diagnosed with COVID, 58% of them took the vaccine. Of the males that were not diagnosed with COVID,  51.5% took the placebo and 48.5% took the vaccine. 
    
### Inferential Results

```{r}
chisq.test(MaleTable)
chisqtestGC(MaleTable)
fisher.test(MaleTable)
```
  
With a small p-value of 0.0005, we can reject the null hypothesis. This indicates that there is a significant difference in COVID infection rate between those those who receive the treatment and those that receive the placebo. The odds of males getting COVID is 0.678 times greater for those that received the placebo rather than the vaccine. 
Overall, this drug is between 35.09% and 45.94% effective for males. This drug is less effective for males than for males and females combined.
  
# Effectivness for Females

## Results

### Graphical Descriptive Results

```{r}
barchartGC(~infected + treatment,data=Females, main= "COVID Infections of females by treatment type", xlab="Infection")
barchartGC(~infected + treatment,data=Females, type="percent", main="Percent COVID Infections of females by treatment type", xlab="Infection")
```
  
Of females that were diagnosed with COVID, about 2 times as many of them received the placebo. Of the females that were not diagnosed with COVID, almost and equal amount received the placebo and the vaccine. 
    
### Numerical Descriptive Results

```{r}
FemaleTable <- xtabs(~infected + treatment, data=Females)
rowPerc(FemaleTable)
colPerc(FemaleTable)
```
  
Of females that were diagnosed with COVID, 67% of them took the placebo and 33% received the vaccine. Of the females that were not diagnosed with COVID,  49.7% received the placebo and 50.3% received the vaccine. 
    
### Inferential Results

```{r}
chisq.test(FemaleTable)
chisqtestGC(FemaleTable)
fisher.test(FemaleTable)
```
  
With a small p-value of 0.00000000000000022, we can reject the null hypothesis. This indicates that there is a significant difference in COVID infection rate between the females who receive the treatment and those that receive the placebo. The odds of females getting COVID is 2.05 times greater for those that received the placebo rather than the vaccine. 
Overall, this drug is between 63.15% and 70.99% effective for females. This drug is more effective for females than for males and females combined.

# Effectivness for Drug Users

## Results

### Graphical Descriptive Results

```{r}
barchartGC(~infected + treatment,data=UserDrugs, main= "COVID Infections of drug users by treatment type", xlab="Infection")
barchartGC(~infected + treatment,data=UserDrugs, type="percent", main="Percent COVID Infections of drug users by treatment type", xlab="Infection")
```
  
Of drug users that were diagnosed with COVID, about 2 times as many of them received the placebo. Of the drug users that were not diagnosed with COVID, a slightly larger amount received the placebo.  
    
### Numerical Descriptive Results

```{r}
DrugTable <- xtabs(~infected + treatment, data=UserDrugs)
rowPerc(DrugTable)
colPerc(DrugTable)
```
  
Of drug users that were diagnosed with COVID, about 33% of them took the vaccine and about 67% them received the placebo. Of the drug users that were not diagnosed with COVID,  52% received the placebo and 48% received the vaccine. 
    
### Inferential Results

```{r}
chisq.test(DrugTable)
chisqtestGC(DrugTable)
fisher.test(DrugTable)
```
  
With a small p-value of 0.002273, we can reject the null hypothesis. This indicates that there is a significant difference in COVID infection rate between drug users who receive the treatment and drug users that receive the placebo. The odds of drug users getting COVID is 1.833 times greater for those that received the placebo rather than the vaccine. 
Overall, this drug is between 54.96% and 73.63% effective for drug users This drug is slightly more effective for drug users than overall population. However, note that the confidence interval for the drug user population is wider than for the overall population.
  
# Effectivness for Gay People

Note: Rather than using the term LGBTQ for this dataset, I've chosen to use the term gay. This is because this data set just includes whether someone identifies as gay or straight, not if they identify as one of the other parts of the LGBTQ acronym.

## Results

### Graphical Descriptive Results

```{r}
barchartGC(~infected + treatment,data=Gay, main= "COVID Infections of gay individuals by treatment type", xlab="Infection")
barchartGC(~infected + treatment,data=Gay, type="percent", main="Percent COVID Infections of gay individuals by treatment type", xlab="Infection")
```
  
Of gay individuals that were diagnosed with COVID, over 2 times as many of them received the vaccine. Of the gay individuals that were not diagnosed with COVID, a slightly larger amount received the placebo.  
    
### Numerical Descriptive Results

```{r}
GayTable <- xtabs(~infected + treatment, data=Gay)
rowPerc(GayTable)
colPerc(GayTable)
```
  
Of gay individuals that were diagnosed with COVID, about 29% of them took the placebo and about 67% them received the vaccine. Of the gay individuals that were not diagnosed with COVID,  about 51% received the placebo and about 49% received the vaccine. 
    
### Inferential Results

```{r}
chisq.test(GayTable)
chisqtestGC(GayTable)
fisher.test(GayTable)
```
  
With a small p-value of 0.000005741, we can reject the null hypothesis. This indicates that there is a significant difference in COVID infection rate between gay individuals who receive the treatment and gay individuals that receive the placebo. The odds of gay individuals getting COVID is 0.4013 times greater for those that received the placebo rather than the vaccine. 
Overall, this drug is only between 20.52% and 37.97% effective for gay individuals. This drug is much less effective for gay individuals than for the overall population.
  
## Conclusion and Discussion

